---
title: "EDA"
author: "Stella Koo"
date: "2024-11-16"
output: github_document
---

## Ian, Stella

## Data Cleaning & Preparing for Analysis
```{r message = FALSE}
library(tidyverse)
library(pheatmap)
library(corrplot)

cleveland = read_csv("./data/cleveland.csv", na = "?")
hungary = read_csv("./data/hungarian.csv", na = "?")
long_beach = read_csv("./data/long_beach_va.csv", na = "?")
switzerland = read_csv("./data/switzerland.csv", na = "?")
```

```{r}
dfs = list(cleveland, hungary, long_beach, switzerland)
regions = c("Cleveland", "Hungary", "Long Beach", "Switzerland")

combined_df = purrr::map2(dfs, regions, ~ mutate(.x, region = .y)) |>
  bind_rows()
```

```{r}
# Number of missing values: should consider NOT analyzing slope, ca, thal
colSums(is.na(combined_df))
```

## Continuous Variables
### Descriptive Statistics
```{r warning = FALSE}
# Summary of continuous variables 
continuous_var = combined_df |>
  group_by(region) |>
  summarise(across(c(age, trestbps, chol, thalach, oldpeak), list(mean = mean, sd = sd, median = median), na.rm = TRUE)) |>
  knitr::kable(digits = 2)

continuous_var
```

### Correlation heatmap for each region
```{r}
cols_to_include = c("age", "trestbps", "chol", "thalach", "oldpeak")
corr_plots = list()

for(r in regions) {
  region_data = combined_df |> filter(region == r)
  
  corr_matrix = cor(region_data[, cols_to_include], use = "complete.obs")
  
  corr_plots[[r]] = corrplot::corrplot(corr_matrix, 
                                       method = "color", 
                                       addCoef.col = "black", 
                                       tl.col = "black", 
                                       main = paste("Correlation Heatmap for", r))
}

```

## Discrete Variables
### Descriptive Statistics
sex, cp, fbs, restecg, exang

* The name and/or source of your data, a description of all relevant variables, and the observational unit (i.e., row). 
Exploratory Data Analysis (EDA):
* Appropriate summary statistics with adequate explanation and interpretation.  	
    * examine binary or other relationships, as well as describing individual distributions.
    * Consider making an appropriate table 
* Graphical displays with adequate explanation and interpretation. 
    * histograms, density plots, and box plots   to visualize distributions of key variables, such as cholesterol, age, thalach (maximum heart rate)
    * create a correlation heatmap to identify relationships between variables and detect multicollinearity, especially for predictors that may be used in regression models.
* Interactive Variables Visualization: Primary Variables:
    * Age vs. Maximum Heart Rate (thalach): Use this to explore how age relates to heart rate, and color points based on heart disease status (num) for additional insight.
    * Age vs. ST Depression (oldpeak): This can reveal if age affects exercise-induced ST depression differently for those with and without heart disease.
* Histograms or Density Plots for Distribution Insights
    * Variables:Age, Cholesterol (chol), num
