---
title: "Exploratory Data Analysis"
author: "Stella Koo (bk2959), Yanhao Shen"
date: "2024-11-16"
output:   
  html_document: default
  github_document: default
---

# Issues to discuss
* Have trouble resizing correlation heatmap

## Data Cleaning & Preparing for Analysis
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(lsr)
library(vcd)

cleveland = read_csv("./data/cleveland.csv", na = "?")
hungary = read_csv("./data/hungarian.csv", na = "?")
long_beach = read_csv("./data/long_beach_va.csv", na = "?")
switzerland = read_csv("./data/switzerland.csv", na = "?")
```

```{r message = FALSE, warning = FALSE}
dfs = list(cleveland, hungary, long_beach, switzerland)
regions = c("Cleveland", "Hungary", "Long Beach", "Switzerland")

combined_df = map2(dfs, regions, ~ mutate(.x, region = .y)) |>
  bind_rows()
```

## Summary of variables 
The tables below displays descriptive statistics for each variable in the heart disease dataset, organized by region. The continuous variables include `age`, `trestbps`, `thalach`, `oldpeak` while the discrete variables consist of `sex`, `cp`, `fbs`, `restecg`, `exang`. These variables will be used to assess the diagnosis of heart disease, as indicated by the `num` variable.

Variables `ca` and `thal` were excluded from the analysis due to incomplete data, with missing values exceeding 85% in more than two regions, rendering these variables non-comparable. Additionally, the `chol` variable was removed from the analysis as all observations in Switzerland had a value of 0

### Cleveland
```{r message = FALSE, warning = FALSE}
summary = function(dataset) {
  dataset_summary = skimr::skim(dataset) |> 
    as.data.frame() |>  
    select(2:6)

  knitr::kable(dataset_summary, digits = 2)
}

summary(cleveland)
```

### Hungary
```{r message = FALSE, warning = FALSE}
summary(hungary)
```

### Long Beach
```{r message = FALSE, warning = FALSE}
summary(long_beach)
```

### Switzerland
```{r message = FALSE, warning = FALSE}
summary(switzerland)
```

## Distribution of participants
The plot illustrates the distribution of participants by age and sex across different regions, with respect to the diagnosis of heart disease. It is evident that the distribution varies significantly by region, likely due to differences in participant availability for the study across these regions. 

Overall, there are noticeably more male participants, and the age distribution is relatively sparse for both groups—those without heart disease and those with heart disease. However, distinct patterns emerge when considering the regional breakdown. To gain a deeper understanding of the factors influencing heart disease prevalence, a more detailed analysis is necessary to identify which variables are contributing to the observed trends, and how these factors may differ or remain consistent across regions.

```{r message = FALSE, warning = FALSE}
ggplot(combined_df, aes(x = age, y = num, color = factor(sex, levels = c(0,1), labels = c("Female", "Male")))) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ region) +
  theme_minimal() +
  labs(title = "Distribution of participants",
       x = "Age",
       y = "Heart Disease Status",
       color = "Sex")
```

In Hungary, the heart disease status (`num`) is limited to values 0 or 1, with no data available for statuses 2, 3, and 4, which also indicate the presence of heart disease. To ensure consistency across regions, the `num` variable was reclassified into a binary format, with 0 representing no heart disease and 1 representing heart disease.

```{r message = FALSE, warning = FALSE}
combined_df = combined_df |>
  mutate(num = factor(ifelse(num == 0, 0, 1), levels = c(0, 1)),
  sex = factor(sex, levels = c(0, 1)),
  slope = factor(slope),
  thal = factor(thal),
  cp = factor(cp),
  fbs = factor(fbs),
  restecg = factor(restecg),
  exang = factor(exang))
```

## Continuous Variables
### Correlation heatmap for each region
A correlation heatmap was generated for each region to examine the relationships between the variables, assess whether these correlations vary across regions, and to anayze any potential multicolinearity.

At a glance, the correlations differ across regions, with some showing stronger relationships than others. The following variables were selected based on the highest correlation values observed in each region:

* `age` and `thalach`: A stronger negative correlation was found in Hungary (-0.45) and Cleveland (-0.39) compared to Switzerland (-0.29) and Long Beach (-0.16).
* `age` and `trestbps`: Long Beach exhibited a stronger positive correlation (0.38) than Cleveland (0.28), Hungary (0.26), and Switzerland (0.24).
* `thalach` and `trestbps`: A stronger negative relationship was observed in Switzerland (-0.23) and Hungary (-0.22) compared to Cleveland (-0.05) and Long Beach (0.00).
* `thalach` and `oldpeak`: A stronger negative correlation was seen in Cleveland (-0.34) and Hungary (-0.32), while Long Beach (0.19) and Switzerland (0.17) displayed weak positive correlations. 

Despite these notable correlations, none exceed 0.7, suggesting that each variable has a unique impact on the outcome (heart disease status). Therefore, multicollinearity is not a concern when modeling a regression.

```{r}
cols_to_include = c("age", "trestbps", "thalach", "oldpeak")

corr_plots = list()

for(r in regions) {
  region_data = combined_df |> filter(region == r)
  
  corr_matrix = cor(region_data[, cols_to_include], use = "complete.obs")
  
  corr_plots[[r]] = corrplot::corrplot(corr_matrix, 
                                       method = "color", 
                                       type = "lower",
                                       main = paste("Correlation Heatmap for", r),
                                       addCoef.col = "black", 
                                       tl.col = "black",
                                       tl.srt = 45)

}

```

#### Relationship Between Variables and Heart Disease Status
Boxplots were generated for each continuous variable in relation to heart disease status (`num`) across different regions. This analysis aimed to visually explore potential relationships between these variables and the binary outcome of heart disease status, while also highlighting regional differences in the distributions of continuous variables. 

##### Age (`age`)
Age is generally higher in heart disease cases across all regions, but the degree of overlap varies. Cleveland and Hungary show more distinct separation, with heart disease cases skewed toward older individuals. Long Beach and Switzerland exhibit greater overlap, suggesting weaker age-based differentiation. Outliers in younger individuals with heart disease are present in Switzerland and Long Beach. 

Age is a strong predictor, especially in Cleveland and Hungary, where separation is more pronounced. Regions with greater overlap (Long Beach, Switzerland) suggest the need for interaction terms. 
```{r message = FALSE, warning = FALSE}
box_plot = function(data, variable) {
  plot = ggplot(data, aes(x = factor(num), y = .data[[variable]], fill = factor(num))) +
      geom_boxplot() +
      theme(legend.position = "none") +
      facet_wrap(~region) 
  
  return(plot) 
}

plot_age = box_plot(combined_df, "age")

plot_age + 
  labs(title = "Boxplot of Age by Heart Disease Status",
       x = "heart disease status",
       y = "age")
```

##### Resting Blood Pressure (in mm Hg, `trestbps`)
Resting blood pressure shows minimal separation between heart disease and non-heart disease groups in most regions. Medians and interquartile ranges are similar across heart disease statuses in Cleveland, Hungary, and Long Beach. Switzerland shows slightly better separation, though the overlap remains significant. Outliers are observed in Cleveland and Long Beach, particularly for non-heart disease cases.

Resting blood pressure may have limited predictive value, as the overlap between the groups is significant across all regions. However, it could still contribute as a secondary predictor or in interaction with other variables. 
```{r message = FALSE, warning = FALSE}
plot_trestbps = box_plot(combined_df, "trestbps")

plot_trestbps + 
  labs(title = "Boxplot of Resting Blood Pressure by Heart Disease Status",
       x = "heart disease status",
       y = "resting blood pressure (in mm Hg)")
```

##### Maximum Heart Rate (`thalach`)
Across all regions, individuals with heart disease generally have lower maximum heart rates, though the degree of separation varies. Cleveland and Hungary show larger differences between the two groups. Long Beach and Switzerland show more overlap, particularly in the upper ranges of heart rates. Cleveland also has more outliers among non-heart disease cases.

Maximum heart rate is a strong predictor of heart disease across most regions, given its consistent trend of lower values in heart disease cases. 
```{r message = FALSE, warning = FALSE}
plot_thalach = box_plot(combined_df, "thalach")

plot_thalach + 
  labs(title = "Boxplot of Maximum Heart Rate by Heart Disease Status",
       x = "heart disease status",
       y = "maximum heart rate")
```

##### T Depression (`oldpeak`)
T depression appears higher in individuals with heart disease (status = 1) across all locations compared to those without (status = 0). Cleveland and Hungary show higher median T depression and larger interquartile ranges for heart disease cases compared to non-cases. Long Beach and Switzerland show smaller differences between heart disease and non-heart disease groups, with generally lower T depression levels. Outliers in Cleveland and Hungary are more pronounced, suggesting regional-specific variability. 

T depression is likely a good predictor, particularly for Cleveland and Hungary, as it shows a clear separation between heart disease and non-heart disease cases. However, its predictive power may vary across regions, necessitating region-specific adjustments. 
```{r message = FALSE, warning = FALSE}
plot_oldpeak = box_plot(combined_df, "oldpeak")

plot_oldpeak + 
  labs(title = "Boxplot of T Depression by Heart Disease Status",
       x = "heart disease status",
       y = "T depression")
```



## Catagorical Variables
the datasets have 7 catagorical Variables: \
sex: 1=male; 0=female\
cp (chest pain type): 1=typical angina; 2=atypical angina; 3=non-anginal pain; 4=asymptomatic \
fbs (fasting blood sugar > 120 mg/dl): 1=true; 0=false \
restecg (resting electrocardiographic results): 0=normal; 1=having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV) \
exang (exercise induced angina): 1=yes; 0=no \
slope (the slope of the peak exercise ST segment): 1=upsloping; 2=flat; 3=downsloping \
thal: 3=normal; 6=fixed defect; 7=reversable defect \
To examine the variables' association with num, I will be using Cramér's V, which is a method based on the Chi-Squared statistic from a contingency table of the two variables. It quantifies the strength of the association, providing a value between 0, indicating no association between the variables and 1, indicating a perfect association. Using this menthod, I hope we can identify key predictors that have strong association with num, and thus focus on variables with higher Cramér's V values in the modeling efforts. This is also a good way in visualization into the relationships within the data.\

now, let's look in to the plot, we can see that thal and cp have the highest Cramér's V (>=0.5 and <1), suggesting that they are the most informative categorical predictors of heart disease in the dataset. Then, exang, slope, and sex also perform moderately good (>= 0.3 and <0.5), saying that they contribute meaningful information and should be included in predictive models. However, at the very beginning, we have note that there is a large gap in the number of male and female participants, so to aviod any bias, I would like to discard sex. Finally, as respect for fbs and restecg, they have limited association and may contribute less to predictive accuracy becasue their Cramér's V are both close to 0, so I will also discard these two variables.
```{r}
# Define variables
variables <- c("sex", "cp", "fbs", "restecg", "exang", "slope", "thal")

# Compute Cramér's V
results <- sapply(variables, function(var) {
  x <- combined_df$num
  y <- combined_df[[var]]
  cramersV(x, y)
})

# Create data frame
association_df <- data.frame(Variable = variables, CramersV = results)

# Plot the results
ggplot(association_df, aes(x = reorder(Variable, CramersV), y = CramersV, fill = CramersV)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Association Between Heart Disease Status and Predictor Variables",
    x = "Predictor Variables",
    y = "Cramér's V",
    fill = "Cramér's V"
  ) +
  theme_minimal()
```

```{r}
ggplot(combined_df, aes(x = as.factor(num), fill = factor(sex, levels = c(0, 1), labels = c("Female", "Male")))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Males and Females by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Proportion",
       fill = "Sex") +
  facet_wrap(~ region, ncol = 2)
```

###cp vs num
```{r}
ggplot(combined_df, aes(x = factor(cp), fill = factor(num))) +
  geom_bar(position = "dodge") +
  facet_wrap(~ region) +
  scale_x_discrete(labels = c("1"="typical angina","2"="atypical angina","3"="non-anginal pain","4"= "asymptomatic"))+
  labs(title = "Chest Pain Type Distribution by Heart Disease Status and Region",
       x = "Chest Pain Type (1-4)",
       y = "Count",
       fill = "Heart Disease Status") +
  theme_minimal()
```

###fbs vs num
```{r}
ggplot(combined_df |> filter(!is.na(fbs)), aes(x = factor(fbs), fill = factor(num))) +
  geom_bar(position = "dodge") +
  facet_wrap(~ region) +
  labs(title = "Fasting Blood Sugar Distribution by Heart Disease Status and Region",
       x = "Fasting Blood Sugar (0=≤120 mg/dl, 1=>120 mg/dl)",
       y = "Count",
       fill = "Heart Disease Status") +
  theme_minimal()
```

###restecg vs num
```{r}
ggplot(combined_df |> filter(!is.na(restecg)), aes(x = factor(restecg), fill = factor(num))) +
  geom_bar(position = "dodge") +
  facet_wrap(~ region) +
  labs(title = "Resting ECG Results by Heart Disease Status and Region",
       x = "Resting ECG Results (0-2)",
       y = "Count",
       fill = "Heart Disease Status") +
  theme_minimal()
```


###exang vs num

```{r}
ggplot(combined_df |> filter(!is.na(exang)), aes(x = factor(exang), fill = factor(num))) +
  geom_bar(position = "dodge") +
  facet_wrap(~ region) +
  labs(title = "Exercise Induced Angina by Heart Disease Status and Region",
       x = "Exercise Induced Angina (0=No, 1=Yes)",
       y = "Count",
       fill = "Heart Disease Status") +
  theme_minimal()
```




