---
title: "EDA"
author: "Stella Koo"
date: "2024-11-16"
output: github_document
---

# Issues to discuss

* For Summary of varaibles, should categorical data be included?
* Cannot create a proper reader friendly table 
* Have trouble resizing correlation heatmap
* Cholesterol variable all zero for Switzerland dataset

* Need to decide which graphs to include/remove

## Data Cleaning & Preparing for Analysis
```{r message = FALSE}
library(tidyverse)
library(pheatmap)
library(corrplot)

cleveland = read_csv("./data/cleveland.csv", na = "?")
hungary = read_csv("./data/hungarian.csv", na = "?")
long_beach = read_csv("./data/long_beach_va.csv", na = "?")
switzerland = read_csv("./data/switzerland.csv", na = "?")
```

```{r}
dfs = list(cleveland, hungary, long_beach, switzerland)
regions = c("Cleveland", "Hungary", "Long Beach", "Switzerland")

combined_df = map2(dfs, regions, ~ mutate(.x, region = .y)) |>
  bind_rows()
```

## Summary of variables 
```{r}
summary_df = function(dataset) {
  map(dataset, ~skimr::skim(.x) |>
        select(2:6) |>
        knitr::kable(digits = 2))
}

summary_df(dfs)
```

## Age and Sex Distribution
```{r}
ggplot(combined_df, aes(x = age)) +
  geom_histogram() +
  facet_wrap(~ region) +
  theme_minimal() +
  labs(title = "Age Distribution by Region",
       x = "Age",
       y = "Count")

ggplot(combined_df, aes(x = factor(sex, levels = c(0,1), labels = c("Female", "Male")))) +
  geom_bar() +
  facet_wrap(~ region)+
  labs(title = "Gender Distribution", x = "Sex", y = "Count")+
  theme_minimal()
```

## Continuous Variables
### Correlation heatmap for each region
```{r fig.width=6, fig.height=4}
cols_to_include = c("age", "trestbps", "chol", "thalach", "oldpeak")

corr_plots = list()

for(r in regions) {
  region_data = combined_df |> filter(region == r)
  
  corr_matrix = cor(region_data[, cols_to_include], use = "complete.obs")
  
  par(mar = c(5, 4, 7, 2))
  
  corr_plots[[r]] = corrplot::corrplot(corr_matrix, 
                                       method = "color", 
                                       addCoef.col = "black", 
                                       tl.col = "black", 
                                       main = paste("Correlation Heatmap for", r),
                                       cex.main = 1.5,  
                                       tl.srt = 45,
                                       cex.axis = 0.8)
}

```

### Bivariate Analysis of Strongly Correlated Variables
Based on the correlation heatmap, we decided to focus on comparing variables that exhibit stronger correlations with each other in each region.

#### `thalach` vs `age`
```{r}
ggplot(combined_df, aes(x = age, y = thalach)) +
  geom_point(aes(color = as.factor(num)), size = 2) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Scatter Plot of Age vs Maximum Heart Rate",
       x = "Age",
       y = "Maximum Heart Rate",
       color = "Heart Disease Status") +
  facet_wrap(~ region, ncol = 2)
```

#### `trestbps` vs `age`
```{r}
ggplot(combined_df, aes(x = age, y = trestbps)) +
  geom_point(aes(color = as.factor(num)), size = 2) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Scatter Plot of Age vs Resting Blood Pressure",
       x = "Age",
       y = "Resting Blood Pressure (in mm Hg)",
       color = "Heart Disease Status") +
  facet_wrap(~ region, ncol = 2) +
  ylim(50, 200)
```

#### `thalach` vs `trestbps`
```{r}
ggplot(combined_df, aes(x = thalach, y = trestbps)) +
  geom_point(aes(color = as.factor(num)), size = 2) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Scatter Plot of Maximum Heart Rate vs Resting Blood Pressure",
       x = "Maximum Heart Rate",
       y = "Resting Blood Pressure (in mm Hg)",
       color = "Heart Disease Status") +
  facet_wrap(~ region, ncol = 2) +
  ylim(50, 200)
```

#### `thalach` vs `oldpeak`
```{r}
ggplot(combined_df, aes(x = thalach, y = oldpeak)) +
  geom_point(aes(color = as.factor(num)), size = 2) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Scatter Plot of Maximum Heart Rate vs T depression",
       x = "Maximum Heart Rate",
       y = "T depression",
       color = "Heart Disease Status") +
  facet_wrap(~ region, ncol = 2) 
```


### Relationship Between Key Variables and Heart Disease Status
Each of the variables explored above was then compared to the Heart Disease Status to assess their relationships and potential impact.

#### `thalach` vs `num`
```{r}
ggplot(combined_df, aes(x = as.factor(num), y = thalach)) +
  geom_boxplot(aes(fill = as.factor(num))) +
  labs(title = "Boxplot of Maximum Heart Rate by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Maximum Heart Rate",
       fill = " Heart Disease Status") +
  facet_wrap(~ region, ncol = 2)
```

#### `trestbps` vs `num`
```{r}
ggplot(combined_df, aes(x = as.factor(num), y = trestbps)) +
  geom_boxplot(aes(fill = as.factor(num))) +
  labs(title = "Boxplot of Resting Blood Pressure by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Resting Blood Pressure (in mm Hg)",
       fill = " Heart Disease Status") +
  facet_wrap(~ region, ncol = 2) +
  ylim(50, 200)
```

#### `oldpeak` vs `num`
```{r}
ggplot(combined_df, aes(x = as.factor(num), y = oldpeak)) +
  geom_boxplot(aes(fill = as.factor(num))) +
  labs(title = "Boxplot of T depression by Heart Disease Status",
       x = "Heart Disease Status",
       y = "T depression",
       fill = " Heart Disease Status") +
  facet_wrap(~ region, ncol = 2) 
```

#### `chol` vs `num`
```{r}
ggplot(combined_df, aes(x = as.factor(num), y = chol)) +
  geom_boxplot(aes(fill = as.factor(num))) +
  labs(title = "Boxplot of Serum Cholestoral by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Serum Cholestoral (mg/dl)",
       fill = " Heart Disease Status") +
  facet_wrap(~ region, ncol = 2) 
```

## Discrete Variables
### Descriptive Statistics
sex, cp, fbs, restecg, exang



