---
title: "Modelling"
author: "Thomas, Yixin Zheng, Yonghao"
date: "2024-11-16"
output: github_document
---

## Thomas, Cecily, Yonghao
```{r setup}
library(tidyverse)
library(janitor)
```

```{r import data author: Yixin Zheng, Stella Koo, include = FALSE}
save_path <- "./data"

column_names <- c("age", "sex", "cp", "trestbps", "chol", "fbs", 
                  "restecg", "thalach", "exang", "oldpeak", "slope", 
                  "ca", "thal", "num")

process_data <- function(file_path, save_name) {

    data <- read.table(file_path, sep = ",", header = FALSE)
    
    colnames(data) <- column_names
    
    write.csv(data, file.path(save_path, save_name), row.names = FALSE)
}

process_data("./heart+disease/processed.cleveland.data", "cleveland.csv")
process_data("./heart+disease/processed.hungarian.data", "hungarian.csv")
process_data("./heart+disease/processed.va.data", "long_beach_va.csv")
process_data("./heart+disease/processed.switzerland.data", "switzerland.csv")
```

```{r data cleaning author: Yixin Zheng}
cleveland <- read_csv("./data/cleveland.csv", na = "?") |> 
  clean_names() # |> drop_na() Removes rows with any missing values (optional, adjust as needed)

hungary = read_csv("./data/hungarian.csv", na = "?") |> 
  clean_names() # |> drop_na() Removes rows with any missing values (optional, adjust as needed)

long_beach = read_csv("./data/long_beach_va.csv", na = "?") |> 
  clean_names() # |> drop_na() Removes rows with any missing values (optional, adjust as needed)

switzerland = read_csv("./data/switzerland.csv", na = "?") |> 
  clean_names() # |> drop_na() Removes rows with any missing values (optional, adjust as needed) 
```
hypothesis:
1. Comparing Diagnostic Factors for Heart Disease Across Regions

Example Hypothesis (SLR):
- **Null Hypothesis (H_0):** There is no linear relationship between cholesterol levels (`chol`) and the presence of heart disease (`num`).
- **Alternative Hypothesis (H_a):** There is a positive linear relationship between cholesterol levels (`chol`) and the presence of heart disease (`num`). ($\beta_1>0$)
  Where:
  - `num` is the target variable indicating heart disease presence.
  - `chol` is the predictor variable representing cholesterol levels.
* there are many null values in the `chol` column in switzerland dataset. So maybe we need to use other diagnostic factors such as `trestbps` if we want to compare the regions

$$ num = \beta_0 + \beta_1 \cdot \text{chol} + \epsilon $$

Example Hypothesis (MLR):
- **Null Hypothesis (H_0):** Cholesterol levels (`chol`), blood pressure (`trestbps`), and exercise-induced angina (`exang`) are not significant predictors of heart disease presence (`num`).
- **Alternative Hypothesis (H_a):** At least one of these variables (`chol`, `trestbps`, `exang`) is a significant predictor of heart disease presence (`num`).

$$ num = \beta_0 + \beta_1 \cdot \text{chol} + \beta_2 \cdot \text{trestbps} + \beta_3 \cdot \text{exang} + \epsilon $$

2. Examining Predictive Power of Clinical Indicators for Heart Disease in Diverse Populations
3. Influence of Age and Lifestyle Factors on Heart Disease
4. Regional Patterns in Heart Disease Diagnostic Attributes