---
title: "Modelling"
author: "Yixin Zheng (yz4993), Thomas Tang (tt3022), Yonghao YU (yy3564)"
date: "2024-11-16"
output: github_document
---

## Yixin Zheng (yz4993), Thomas Tang (tt3022), Yonghao YU (yy3564)
```{r setup}
library(tidyverse)
library(janitor)
```

```{r import data author: Yixin Zheng, Stella Koo, include = FALSE}
save_path <- "./data"

column_names <- c("age", "sex", "cp", "trestbps", "chol", "fbs", 
                  "restecg", "thalach", "exang", "oldpeak", "slope", 
                  "ca", "thal", "num")

process_data <- function(file_path, save_name) {

    data <- read.table(file_path, sep = ",", header = FALSE)
    
    colnames(data) <- column_names
    
    write.csv(data, file.path(save_path, save_name), row.names = FALSE)
}

process_data("./heart+disease/processed.cleveland.data", "cleveland.csv")
process_data("./heart+disease/processed.hungarian.data", "hungarian.csv")
process_data("./heart+disease/processed.va.data", "long_beach_va.csv")
process_data("./heart+disease/processed.switzerland.data", "switzerland.csv")
```


Interpreting num (author: Yixin Zheng):
The values for `num` represent the degree of narrowing in the coronary arteries:
0: No disease (< 50% diameter narrowing).
1-4: Increasing severity of disease (> 50% diameter narrowing, with different severities).

For convenience, this variable will binarized:
0: No heart disease (value 0 in num).
1: Presence of heart disease (values 1-4 in num).

*but if we want to analyze the severity of heart disease num will be treated as a categorical variable.
example code:
cleaned_data <- data |> 
  mutate(num = factor(num, levels = c(0, 1, 2, 3, 4), 
                      labels = c("No Disease", "Mild", "Moderate", "Severe", "Very Severe")))
                      
```{r data cleaning author: Yixin Zheng}
cleveland <- read_csv("./data/cleveland.csv", na = "?") |> 
  clean_names() |> 
  mutate(num = if_else(num == 0, 0, 1)) # Binarize the `num` variable: 0 = no heart disease, 1 = heart disease

# |> drop_na() Removes rows with any missing values (optional, adjust as needed)

hungary = read_csv("./data/hungarian.csv", na = "?") |> 
  clean_names() |> 
  mutate(num = if_else(num == 0, 0, 1))
# |> drop_na() Removes rows with any missing values (optional, adjust as needed)

long_beach = read_csv("./data/long_beach_va.csv", na = "?") |> 
  clean_names() |> 
  mutate(num = if_else(num == 0, 0, 1))
# |> drop_na() Removes rows with any missing values (optional, adjust as needed)

switzerland = read_csv("./data/switzerland.csv", na = "?") |> 
  clean_names() |> 
  mutate(num = if_else(num == 0, 0, 1))
# |> drop_na() Removes rows with any missing values (optional, adjust as needed) 

```

```{r}
cor(cleveland$chol, cleveland$num, use = "complete.obs")

```


hypothesis (author: Yixin Zheng)
* need to run some test choose explanatory variables?

1. Comparing Diagnostic Factors for Heart Disease Across Regions
 explore whether certain diagnostic factors (e.g., cholesterol levels, exercise-induced angina) are more predictive of heart disease in one region compared to others.
 
* there are many null values in the `chol` column in switzerland dataset. So maybe we need to use other diagnostic factors such as `trestbps` if we want to compare the regions
Example Hypothesis (SLR):
- **Null Hypothesis (H_0):** There is no linear relationship between cholesterol levels (`chol`) and the presence of heart disease (`num`).
- **Alternative Hypothesis (H_a):** There is a positive linear relationship between cholesterol levels (`chol`) and the presence of heart disease (`num`). ($\beta_1>0$)
  Where:
  - `num` is the target variable indicating heart disease presence.
  - `chol` is the predictor variable representing cholesterol levels.

$$ num = \beta_0 + \beta_1 \cdot \text{chol} + \epsilon $$

Example Hypothesis (MLR):
- **Null Hypothesis (H_0):** Cholesterol levels (`chol`), blood pressure (`trestbps`), and exercise-induced angina (`exang`) are not significant predictors of heart disease presence (`num`).
- **Alternative Hypothesis (H_a):** At least one of these variables (`chol`, `trestbps`, `exang`) is a significant predictor of heart disease presence (`num`).

$$ num = \beta_0 + \beta_1 \cdot \text{chol} + \beta_2 \cdot \text{trestbps} + \beta_3 \cdot \text{exang} + \epsilon $$

2. Examining Predictive Power of Clinical Indicators for Heart Disease in Diverse Populations

Example Hypothesis (SLR):
- **Null Hypothesis (H_0):** There is no linear relationship between maximum heart rate achieved (`thalach`) and the presence of heart disease (`num`).
- **Alternative Hypothesis (H_a):** There is a negative linear relationship between maximum heart rate achieved (`thalach`) and the presence of heart disease (`num`). ($\beta_1<0$)

$$ num = \beta_0 + \beta_1 \cdot \text{thalach} + \epsilon $$
Example Hypothesis (MLR):
- **Null Hypothesis (H_0):** Maximum heart rate (`thalach`), fasting blood sugar (`fbs`), and the slope of the ST segment (`slope`) are not significant predictors of heart disease presence (`num`).
- **Alternative Hypothesis (H_a):** At least one of these variables (`thalach`, `fbs`, `slope`) is a significant predictor of heart disease presence (`num`).

$$ num = \beta_0 + \beta_1 \cdot \text{thalach} + \beta_2 \cdot \text{fbs} + \beta_3 \cdot \text{slope} + \epsilon $$

3. Influence of Age and Lifestyle Factors on Heart Disease

Example Hypothesis (SLR):
- **Null Hypothesis (H_0):** There is no linear relationship between age (`age`) and the presence of heart disease (`num`).
- **Alternative Hypothesis (H_a):** There is a positive linear relationship between age (`age`) and the presence of heart disease (`num`). ($\beta_1>0$)

$$ num = \beta_0 + \beta_1 \cdot \text{age} + \epsilon $$
Example Hypothesis (MLR):
- **Null Hypothesis (H_0):** Age (`age`), chest pain type (`cp`), and exercise-induced angina (`exang`) are not significant predictors of heart disease presence (`num`).
- **Alternative Hypothesis (H_a):** At least one of these variables (`age`, `cp`, `exang`) is a significant predictor of heart disease presence (`num`).

$$ num = \beta_0 + \beta_1 \cdot \text{age} + \beta_2 \cdot \text{cp} + \beta_3 \cdot \text{exang} + \epsilon $$

4. Regional Patterns in Heart Disease Diagnostic Attributes

Example Hypothesis (SLR):
- **Null Hypothesis (H_0):** There is no linear relationship between diagnostic attributes (e.g., cholesterol, age) and heart disease (`num`) in each region.
- **Alternative Hypothesis (H_a):**The relationship between diagnostic attributes (e.g., cholesterol, age) and heart disease (`num`) differs significantly across regions.
- This hypothesis can be tested by performing SLR for each region and comparing the coefficients (beta_1 values) to see if they vary.

Example Hypothesis (MLR):
- **Null Hypothesis (H_0):** Regional differences do not significantly influence the predictive power of diagnostic factors for heart disease.
- **Alternative Hypothesis (H_a):** Regional differences significantly influence the predictive power of diagnostic factors for heart disease.

$$ num = \beta_0 + \beta_1 \cdot \text{chol} + \beta_2 \cdot \text{age} + \beta_3 \cdot \text{region} + \beta_4 \cdot (\text{chol} \times \text{region}) + \beta_5 \cdot (\text{age} \times \text{region}) + \epsilon $$


author: Thomas Tang
```{r }
cleveland <- read.csv("./data/cleveland.csv", header = FALSE)
hungarian <- read.csv("./data/hungarian.csv", header = FALSE)
long_beach <- read.csv("./data/long_beach_va.csv", header = FALSE)
switzerland <- read.csv("./data/switzerland.csv", header = FALSE)

# Add region column and combine datasets
cleveland$region <- "Cleveland"
hungarian$region <- "Hungarian"
long_beach$region <- "Long_Beach_VA"
switzerland$region <- "Switzerland"
combined_data <- bind_rows(cleveland, hungarian, long_beach, switzerland)

colnames(combined_data) <- c("age", "sex", "cp", "trestbps", "chol", "fbs",
                             "restecg", "thalach", "exang", "oldpeak", "slope",
                             "ca", "thal", "num", "region")
combined_data <- combined_data %>%
  mutate(across(c(age, sex, cp, trestbps, chol, fbs, restecg, thalach,
                  exang, oldpeak, slope, ca, thal, num), as.numeric))
```

```{R}
cleaned_data <- combined_data %>% select(-ca, -thal)
critical_columns <- c("num", "age", "sex", "cp", "trestbps", "chol",
                      "fbs", "restecg", "thalach", "exang", "oldpeak", "slope")
cleaned_data <- cleaned_data %>% drop_na(all_of(critical_columns))
summary(cleaned_data)
```
```{r author: Thomas Tang}
cleaned_data$num <- ifelse(cleaned_data$num > 0, 1, 0)
cleaned_data$num <- as.factor(cleaned_data$num)
logistic_model <- glm(num ~ age + sex + cp + trestbps + chol + fbs +
                      restecg + thalach + exang + oldpeak,
                      data = cleaned_data, family = binomial)
summary(logistic_model)
```
(author= Thomas Tang)
Significant Predictors (p-value < 0.05):
sex (1.409350):
Males are significantly more likely to have heart disease than females.
Odds ratio: exp(1.409)=4.09 → Males are 4.09 times more likely to have heart disease compared to females.
cp (0.687260):
Higher chest pain levels increase the odds of heart disease.
Odds ratio: exp(0.687)=1.99.
trestbps (0.013883):
Resting blood pressure has a small but significant positive effect.
Odds ratio: exp(0.0139)=1.014 per unit increase.
chol (-0.003090):
Cholesterol has a small negative association.
Odds ratio: exp(−0.0031)=0.997, slightly decreasing odds.
thalach (-0.021518):
Higher maximum heart rate achieved decreases the odds of heart disease.
Odds ratio: exp(−0.0215)=0.978 → Protective effect.
exang (0.894064):
Exercise-induced angina increases the odds of heart disease. exp(0.894)=2.44.
oldpeak (0.586433):Higher ST depression values significantly increase the odds of heart disease.
Odds ratio: exp(0.586)=1.80.
fbs (Fasting blood sugar) and age do not have significant association with heart disease in this dataset.








